Data Formatting

### 2021-02-19 scripts/data_format.Rmd
cleaning up code

Packages
```{r}
library(tidyverse)
library(magrittr)
library(vegan)

library(here)
```

@@

@@

@@

community matrix for PCA - primarily used by fig2
```{r}
#select for community variables
community_mat <- data %>% dplyr::select(WetAltID, Month.1, AB2:AB8, AB9, AB20:AB42) 
community_mat$Month.1 <- gsub("Month", "", community_mat$Month.1)
community_mat %<>% mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>% distinct() %>% arrange(.,WetAltID) %>% dplyr::select(-WetAltID, -Month.1) %>% dplyr::select(siteID, AB2:AB42)
#rownames
community_mat %<>% column_to_rownames(., var = "siteID")

#select for environmental variables
env <- data %>% dplyr::select(WetAltID, Month.1, MeanAirT, MeanWaterTempPredC, DryingScore, CanopyCover, Area, Perimeter) %>% distinct() %>% arrange(.,WetAltID) 
env$Month.1 <- gsub("Month", "", env$Month.1)
#make siteID rowname
env$siteID <- paste(env$WetAltID, env$Month.1, sep = "_")
env %<>% dplyr::select(-WetAltID, -Month.1)
#remove duplicated rows (siteID is duplicated but env. variables are not... not sure why this is)
env <- env[-c(53, 90),]
env_mat <- env %>% column_to_rownames(., var = "siteID")

#replace NA's with zeros. Probably not the best option but will allow analysis to run for now
community_mat[is.na(community_mat)] <- 0
env_mat[is.na(env_mat)] <- 0

#pca on community matrix and gather scores for first two principal components
pca <- princomp(community_mat, scores = TRUE)
summary(pca)
scores <- as.data.frame(pca$scores)
scores %<>% dplyr::select(Comp.1, Comp.2) %>% rownames_to_column(., var = "siteID")

#plot pca
biplot(pca)

tmp <- data %>% dplyr::select(WetAltID, Month.1, cc, Month, AB2:AB42) %>% 
  dplyr::select(-ABAmb, -ABSal) %>%  mutate(., Size = rowSums(.[5:25])) %>% #get totals for community size
  dplyr::select(-AB2:-AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>%
  distinct()
tmp$siteID <- gsub("Month", "", tmp$siteID)

site_scores  <- inner_join(tmp, scores, by = "siteID")

@@

cc x evenness data - primarily used by fig3
```{r}
#Get species richness from matrix

community_mat <- data %>% dplyr::select(WetAltID, Month.1, AB2:AB8, AB9, AB20:AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>% distinct()

community_mat <- community_mat[order(community_mat$WetAltID, community_mat$Month.1),]

abundances <- community_mat

community_mat %<>% column_to_rownames(., var = "siteID")

presence_mat <- ifelse(community_mat>0, 1, 0)

richness_mat <- tibble()
for (i in 1:length(siteID)){
  richness_mat[i,1] <- siteID[i]
  richness_mat[i,2] <- sum(presence_mat[i,])
}
richness_mat %<>% rename(siteID = 1, richness = 2)

order <- richness_mat

richness_cc <- full_join(tmp, richness_mat, by = "siteID")

@@

h <- vegan::diversity(community_mat, index = "shannon")
h <- melt(as.matrix(h))
h %<>% rename(siteID = Var1, h = value) %>% dplyr::select(-Var2)

@@

evenness %<>% full_join(h, evenness, by = "siteID")

evenness %<>% mutate(J = h/lnS) %>% rowwise()

#create new df for ordered sites to test lagged richness and evenness on prevalence
lag_evenness <- evenness[-c(4,14,47,90),]
lag_evenness$lag_richness <- c(0)
lag_evenness$lag_J <- c(0)
lag_evenness$lag_cc <- c(0)
lag_evenness$lag_size <- c(0)
for(n in 2:91){
  lag_evenness$lag_richness[n] <- lag_evenness$richness[n-1]
  lag_evenness$lag_J[n] <- lag_evenness$J[n-1]
  lag_evenness$lag_cc[n] <- lag_evenness$cc[n-1]
  lag_evenness$lag_size[n] <- lag_evenness$size[n-1]
}

@@

#fit two lines to cc_evenness_prev plot
lag_evenness$high <- c(0)
lag_evenness$low <- c(0)
for (i in 1:nrow(lag_evenness)){
  if (lag_evenness$lag_J[i] > 0.6){
    lag_evenness$high[i] <- 1
    lag_evenness$low[i] <- 1
  }
  else if (lag_evenness$lag_cc[i] < 45000){
    lag_evenness$low[i] <- 1
  }
  else if (lag_evenness$lag_cc[i] > 45000){
    lag_evenness$high[i] <- 1
  }
  else {
    lag_evenness$high[i] <- 0
    lag_evenness$low[i] <- 0
  }
}
low_cc <- lag_evenness %>% filter(low==1)
cor.test(x=low_cc$lag_J, y=low_cc$lag_cc,method = "spearman")
high_cc <- lag_evenness %>% filter(high==1)
cor.test(x=high_cc$lag_J, y=high_cc$lag_cc,method = "spearman")
```

>>>>

Packages
```{r}
library(tidyverse)
library(magrittr)
library(vegan)
library(reshape)
library(here)
```

@@

community summary - includes each site_month as rows and variables cc, size, prevalenceetc... as columns
```{r}
comm_summ <- data %>% dplyr::select(WetAltID, Month.1, cc, Month, AB2:AB42, Prevalence) %>% 
  dplyr::select(-ABAmb, -ABSal) %>%  mutate(., Size = rowSums(.[5:25])) %>% #get totals for community size
  dplyr::select(-AB2:-AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>%
  distinct()
comm_summ$siteID <- gsub("Month", "", comm_summ$siteID)
```

@@

community matrices - abundance matrix and presence-absence matrix
```{r}
community_mat <- data %>% dplyr::select(WetAltID, Month.1, AB2:AB8, AB9, AB20:AB42) 
community_mat$Month <- gsub("Month", "", community_mat$Month.1)
community_mat %<>% mutate(., siteID = paste(WetAltID, Month, sep = "_")) %>% distinct() %>% arrange(.,WetAltID) %>% dplyr::select(siteID, AB2:AB42)

#make abundance and pres/abs matrices with rownames as siteIDs
abundance_mat <- community_mat %>% column_to_rownames(., var = "siteID")
presence_mat <- as.data.frame(ifelse(abundance_mat>0, 1, 0))
```

environmental variables matrix
```{r}
#select for environmental variables
env <- data %>% dplyr::select(WetAltID, Month.1, MeanAirT, MeanWaterTempPredC, DryingScore, CanopyCover, Area, Perimeter) %>% distinct() %>% arrange(.,WetAltID) 
env$Month <- gsub("Month", "", env$Month.1)
#make siteID rowname
env %<>% mutate(., siteID = paste(env$WetAltID, env$Month, sep = "_")) %>% distinct() %>% arrange(.,WetAltID) %>% dplyr::select(siteID, MeanAirT:Perimeter)

#need to remove duplicated rows and determine what to do with NA's
env_mat <- env %>% column_to_rownames(., var = "siteID")
```
environmental variables are `r colnames(env_mat)`

site 13, month 1 and site 19, month 4 are duplicated. not sure why

richness values for each site
```{r}
#Get species richness from pres/abs matrix
richness <- tibble()
for (i in 1:length(rownames(presence_mat))){
  richness[i,1] <- rownames(presence_mat)[i]
  richness[i,2] <- sum(presence_mat[i,])
}
richness %<>% rename(siteID = 1, richness = 2)
```

@@

PCA for community matrix - primarily used by fig2
```{r}
#pca on community matrix and gather scores for first two principal components
pca <- princomp(abundance_mat, scores = TRUE)
summary(pca)
pca_scores <- as.data.frame(pca$scores)
pca_scores %<>% dplyr::select(Comp.1, Comp.2) %>% rownames_to_column(., var = "siteID")

site_scores  <- full_join(comm_summ, pca_scores, by = "siteID")

@@

cc x evenness data - primarily used by fig3
```{r}
richness_cc <- full_join(comm_summ, richness_mat, by = "siteID")

@@

shannon <- vegan::diversity(abundance_mat, index = "shannon")
shannon <- melt(as.matrix(shannon))
shannon %<>% mutate(siteID = X1, h = value) %>% dplyr::select(siteID, h)

@@

evenness %<>% full_join(shannon, evenness, by = "siteID")

evenness %<>% mutate(J = h/lnS) %>% rowwise()
evenness <- evenness[order(evenness$WetAltID),]
#create new df for ordered sites to test lagged richness and evenness on prevalence
lag_evenness <- evenness[-c(4,14,47,90),] #brute force remove site_months out of sequence
lag_evenness$lag_richness <- c(0)
lag_evenness$lag_J <- c(0)
lag_evenness$lag_cc <- c(0)
lag_evenness$lag_size <- c(0)
for(n in 2:91){
  lag_evenness$lag_richness[n] <- lag_evenness$richness[n-1]
  lag_evenness$lag_J[n] <- lag_evenness$J[n-1]
  lag_evenness$lag_cc[n] <- lag_evenness$cc[n-1]
  lag_evenness$lag_size[n] <- lag_evenness$size[n-1]
}

@@
